---
title: 'README'
author: 'Evan Brown'
date: '8/1/2022'
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r global options, include = FALSE}
knitr::opts_chunk$set(fig.path = "README_files/figure-gfm")
```

### Brief description of the app and its purpose.  
 
I created this shiny app for the NCSU ST558 Summer '22 Final Project. This app is a culmination of real data collected by me as part of my Research Toxicology Fellowship at the US Environmental Protection Agency. I hope this app is able to give you insight into the new predictive risk assessment approach the EPA has been implementing for the past few years. In 2020, I was put on a team to develop a new assay to screen chemicals for androgen receptor (AR) toxicity. In late 2021, the assay was developed and I ran a validation screen of 128 chemicals (plus 2 control chemicals). Each chemical was evaluated with 11 cytochrome p450 enzymes (plus 2 control enzymes), then assay endpoints were measured (AR activity and cell viability). Each chemical was run in dose-response, and the final product was a large, complex study data set. Hit calls in the AR assay were made and distinguished from purely cytotoxic chemicals. Then, each chemical was run through a chemotyping software, developing a second large data set containing structural predictors for each chemical. Finally, these two data sets were combined into a chemotype-hitcall data set that provides a connection between structural chemical features and assay hits. It's this chemotype-hitcall data set that is crucial to the modeling and prediction approach demonstrated in this app.

In the About page, I'll introduce you to the assay, give background, and provide links for further information. In the Data Exploration page, you'll be able to filter the raw study data before being mapped to hit calls by chemical, biogroup, and both. Then, you'll be able to save your filtered data set to your local machine. In the Data Visualization page, you'll be able to interact with hill-curve scatter plots of the data. Selecting different cyp's will allow you to visualize the role of liver metabolism in breaking down drugs inside your body. In particular, the chemical Flutamide was able to be shifted dramatically by two cyps. Hit calls were made via a series of criteria based on AR and viability data and final call is displayed on the AR plots. In the modeling page, you'll be able to visualize the full chemotype-hitcall data set, the chemotype-hitcall data set with only present chemotypes included, and a low-dimensional version used for modeling the generalized linear model. Then, you'll be able to fit three models, with the Random Forest model frequently outperforming the rest. Finally and most exciting, you'll be able to predict androgen receptor hit calls of untested chemicals, view chemical descriptions and even have a look at chemical structures. Ideally, we'd be able to run these chemicals back through the assay to determine how well our model predicted these, but for now we'll settle for the prediction. 

In a nutshell, assay data was collected, chemicals were broken down into their structural components, then those components were mapped to the hit call results in the assay to train models to predict assay hits based on chemical features. The ultimate goal of the project and of the new EPA approach simulated here was to be able to predict toxicity for untested chemicals based on chemical structure. When we were tasked with this project, I immediately thought of the idea to do it this way and couldn't let it go. I had discussions with several folks from the lab and from higher up to get the green light to tackle this, and I am happy with how it's turned out. I've been asked to present this work mid-August. However, while completing the project, I realized some things were not able to meet the rubric requirements perfectly. I took confidence in the open-endedness mentioned there and I hope you'll see the tremendous amount of thought and effort that went in to both trying to meet those requirements while also making this app reflective of real work taking place that will have an impact on chemical regulatory decisions in the future.

### A list of packages needed to run the app.  

-shiny  
-ggplot2  
-caret  
-randomForest  
-gbm  
-tidyverse  
-data.table  
-shinythemes  
-mathjaxr  

### A line of code that would install all the packages used.  

```{r, eval= FALSE}
install.packages("shiny", "ggplot2", "caret", "randomForest", "gbm", "tidyverse", "data.table", "shinythemes", "mathjaxr")
library(shiny)
library(ggplot2)
library(caret)
library(randomForest)
library(gbm)
library(tidyverse)
library(data.table)
library(shinythemes)
library(mathjaxr)
```

### Code to run the app.

```{r, eval = FALSE}
shiny::runGitHub(repo="Shiny-Project",username = "ecbrown7",ref = "main",subdir = "AR-Shiny-App")
```
